A significance test for forward stepwise model selection

### 2014-05-12 paper/paper.tex

With non-trivial groups, however, the event that $g^*(y)$, the first
group chosen, is equal to some fixed group $g$ is generally not given by
a set of affine constraints. Rather, considering line 3 of Algorithm~\ref{algo:fs}, we see
\begin{equation}
\label{eq:first:quadratic}
\begin{aligned}
\{g^*(y)=g\} &= \{ \|X_g^Ty\|_2 /w_g \geq \|X_h^Ty\|_2 /w_h , \ \forall h \neq g\} \\
&=  \{ y^T(X_g^TX_g)y / w_g^2 - y^T(X_h^TX_h)y / w_h^2 \geq 0, \ \forall h \neq g\} \\
\end{aligned}
\end{equation}
Hence, our selection event can be expressed as 
the intersection of a list of quadratic inequalities. This non-affine selection event is stylized in Figure
\ref{fig:curved}.
In the rest of this section, we consider an arbitrary selection procedure
$S$ that returns one of a set of possible outcomes $s \in {\cal S}$ by a set of quadratic inequalities.
That is, for all possible outcomes $s \in {\cal S}$ there is a set of indices
$I(s)$ such that
\begin{equation}
\label{eq:selection}
\begin{aligned}
\{S(y)=s\} = \cap_{i \in I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
\end{equation}
The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.
As in \cite{lasso:fixed} we will be interested in some randomly chosen contrast
$\eta(S(y))^T\mu$.
In our grouped selection procedure, the selection rule is $S(y)=g^*(y)$
and $\eta$ is typically 
$$
\eta(S(y)) = \eta(g^*(y)) = \frac{X_{g^*(y)}^Ty}{\|X_{g^*(y)}^Ty\|_2}.
$$
This choice of $\eta$ reflects our goal of calculating a p-value for inclusion of the variable $g^*$ chosen by forward stepwise.

@@

@@

Lemma \ref{eq:test:dbn} gives an explicit characterization of our p-value
that can also be used to calculate it in practice. An implementation only needs to compute the intersection of all the intervals implied by each quadratic constraint \eqref{eq:quadslice} and then evaluate the $\chi$ distribution truncated to the intersection of these intervals at the observed value.
For an alternative
derivation including a detailed algorithm description for the implementation used throughout this paper see Appendix~\ref{app:algo}.

After adding the first group in forward stepwise and computing its p-value, we orthogonalize the response $y$ and all remaining groups. This imposes some constraints on $y$ that in principle should be used when computing p-values at subsequent steps. For now we ignore these constraints and iterate the global null test, but work on incorporating all the known constraints is ongoing. We believe that the effect of not tracking all constraints is what causes the nominal p-value to be increasingly stochastically larger than uniform further down the forward stepwise path, and only truly uniform at the first step where all remaining variables are noise.

>>>>

With non-trivial groups, however, the event that $g^*(y)$, the first
group chosen, is equal to some fixed group $g$ is generally not given by
a set of affine constraints. Rather, considering line 3 of Algorithm~\ref{algo:fs}, we see
\begin{equation}
\label{eq:first:quadratic}
\begin{aligned}
\{g^*(y)=g\} &= \{ \|X_g^Ty\|_2 /w_g \geq \|X_h^Ty\|_2 /w_h , \ \forall h \neq g\} \\
&=  \{ y^T(X_g^TX_g)y / w_g^2 - y^T(X_h^TX_h)y / w_h^2 \geq 0, \ \forall h \neq g\} \\
\end{aligned}
\end{equation}
Hence, our selection event can be expressed as 
the intersection of a list of quadratic inequalities. This non-affine selection event is stylized in Figure
\ref{fig:curved}.
In the rest of this section, we consider an arbitrary selection procedure
$S$ that returns one of a set of possible outcomes $s \in {\cal S}$ by a set of quadratic inequalities.
That is, for all possible outcomes $s \in {\cal S}$ there is a set of indices
$I(s)$ such that
\begin{equation}
\label{eq:selection}
\begin{aligned}
\{S(y)=s\} = \cap_{i \in I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
\end{equation}
The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.
As in \cite{lasso:fixed} we will be interested in some randomly chosen contrast
$\eta(S(y))^T\mu$.
In our grouped selection procedure, the selection rule is $S(y)=g^*(y)$
and $\eta$ is typically 
$$
\eta(S(y)) = \eta(g^*(y)) = \frac{X_{g^*(y)}^Ty}{\|X_{g^*(y)}^Ty\|_2}.
$$
Note that for each fixed $g$ and $\eta \in \real^g$ with $\|\eta\|_2=1$, the event 
$$
\left\{y:\frac{X_g^Ty}{\|X_g^Ty\|_2} = \eta \right\} = \left\{y: \|X_g^Ty\|^2_2 \leq (X_g\eta)^Ty \right\},
$$
is defined by another quadratic inequality that can be appended to $I(s)$.

@@

We now present an explicit algorithm for computing this slice in quadratic selection procedures.

\begin{algorithm}
 \caption{Truncation interval for quadratic decisions}
 \label{algo:quadratic}
 \begin{algorithmic}
   \REQUIRE Response $y$, state $s$ list of quadratic inequalities $\{Q_i, a_i, b_i: i \in I(s)\}$, direction
   of interest $\eta$ with $\|\eta\|_2=1$.
   \ENSURE The set $\{t: (y+t \cdot \eta)^TQ_i(y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i  \ \forall i \in I(s) \}$.
      \STATE Initialize  interval: $M= (-\infty,\infty)$, $U=-\infty, L=\infty$
    \FOR{$i$ in $I(s)$}
    \STATE{$a=\eta^TQ_i\eta, b=2y^TQ_i\eta + a_i^T\eta, c=y^TQ_iy + a_i^Ty - b_i$}
    \IF{$a \neq 0$}
    \IF{$b^2-4ac > 0$}
    \STATE Stop: $y$ does not satisfy inequalities!
    \ELSIF{$a > 0$}
     \STATE $M \gets M \cap [(-b-\sqrt{b^2-4ac})/2a, (-b+\sqrt{b^2-4ac})/2a]$;
     \ELSIF{$a < 0$}
     \STATE $L \gets \min(L,(-b-\sqrt{b^2-4ac})/2a)$,
     \STATE $U \gets \max(U,(-b+\sqrt{b^2-4ac})/2a).$;
    \ENDIF
    \ELSE
    \IF{$b > 0$}
   \STATE $M \gets M \cap (-\infty, -c/b]$;
   \ELSIF{$b < 0$}
   \STATE $M \gets M \cap [-c/b,\infty)$;
   \ELSIF{$c > 0$}
   \STATE Stop: $y$ does not satisfy inequalities!
   \ENDIF
   \ENDIF
   \ENDFOR
   \RETURN $(M \cap (L,U)^c) + \eta^Ty.$
 \end{algorithmic}
\end{algorithm}

@@

\subsubsection{The $T\chi$ test statistic}

This leads us to what we call the {\em  $T\chi$ test statistic} which is determined by
a degrees of freedom parameter $r$ as well as a truncation set
$M \subset \real$, a scale parameter $\theta$ and an observed value $t$.
The test statistic is just the survival function of a $\chi_r$ random variable
truncated to the set $M$ evaluated at our observed 
norm $\|X_g^Ty\|_2$ (after scaling by the appropriate scale $\theta$).
\begin{equation}
\label{eq:tchi}
T\chi(t, r, \theta,  M) = \frac{\int_{M /\theta \cap [t/\theta,\infty)} F_{\chi_r}(dz)}{\int_{M / \theta} F_{\chi_r}(dz)}.
\end{equation}
